A MarineOmics overview of methods to detect signatures of selection

### 2025-04-15 DRAFT_Signatures_Selection/DRAFT_Signatures_of_Selection.Rmd
updates to Sig Sel page

---
title: "A MarineOmics overview of methods to detect signatures of selection"
author: "Joanna Griffiths, Danielle Davenport, Madison Armstrong, Camille Rumberger"
date: "2025-02-24"
output:
  html_document:
    toc: yes
bibliography: "common-bib_01.bib"
---

*Created using [HackMD](https://https://hackmd.io/) and [tablesgenerator](https://www.tablesgenerator.com/markdown_tables)*

## Target audience
This page is designed for beginners and provides a broad introduction to different methods commonly used to study natural selection in wild populations at a microevolutionary scale. On this page we will explore how techniques like genotype-environment analysis (GEA) and genome-wide association studies (GWAS) can help us study natural selection. To learn more about specific methods, read through the suggested readings in each section and visit the tutorials where links are provided.

@@

<ins>On this page we describe three classes of methods to identify loci under selection at a microevolutionary scale using samples collected from a single time point across space/environments. .</ins>
1. **Outlier Analyses** identify extreme deviations from what would be expected under neutral evolution. 
2. **Genotype-Environment-Analysis (GEA)** links genomic variation to environmental gradients.
3. **Genome-Wide Association Studies (GWAS)** examine genetic variants and identify those statistically associated with a particular trait. When a specific phenotype shows a significant association with a particular loci, gene or region, it suggests that this genetic variation may be adaptive.

## Which method should I use to find loci under selection?
Briefly, the biology of your system is always important to consider before conducting specific analyses, and no matter which method(s) you use, it is recommended that comparisons of outlier vs non-outliers (neutral loci) are performed [(@riginos2016navigating)](#refs) as a kind-of “validation” step, where we would expect loci under selection (i.e. outliers) to demonstrate a stronger signal of environmental dependence than non-outliers. However, a common but conservative approach taken by researchers performing outlier detection is to apply multiple test methods and identify outlier SNPs that are overlapping across them. For example, some studies perform a differentiation outlier test first (see [Outlier Analyses Section](#outlier-analyses)) and then use only the resulting outliers as input for the Genome-Environment Associations  [(@rellstab2015practical)](#refs).

Importantly, to conduct outlier analyses you do not necessarily need to have neutral genetic differentiation (genetic variation among populations  primarily driven by random processes like genetic drift and migration) in your data. For example, marine invertebrates with pelagic larval dispersal and high gene flow which show very low to no detectable population structure at neutral alleles across their range (i.e using STRUCTURE, ADMIXTURE, PCA) still show signals of adaptation. Other examples include Pacific purple sea urchins ([(@petak2023whole-cp)](#refs);[(@rumberger2023selection)](#refs)) or scallops [(@van2017identifying-cp)](#refs).

@@

Genetic differences among groups may be driven by a combination of genetic drift, demography and/ or selection. Each of these processes can directly change the allele frequency of a population. However, alleles that are under selection and influence fitness are expected to occur at  frequencies different to those affected only by neutral processes. For example, loci under divergent selection (and those physically linked to them) exhibit stronger differentiation than neutral regions and so we might expect these loci to be more differentiated between populations/groups.  We can identify these differences in allele frequencies among different populations/environments using outlier tests. Outlier tests for selection assume a null model of neutrality, where allele frequencies drift randomly without selection pressures. Loci with significantly higher or lower differentiation (i.e. Fst) between populations/groups than expected under a neutral model are flagged as potential targets of selection. Specifically, while selectively neutral loci are expected to form a normal frequency distribution,  when alleles increase fitness (positive selection) they can be expected to be observed at higher frequencies and where they decrease fitness (negative selection) they can be expected to be observed at at lower frequencies [(@hoban2021genetic-cp)](#refs).  These are  “outliers” or “putatively adaptive loci” which may be candidates of selection.

There are two commonly applied approaches to outlier analysis; FST-based analysis (Figure 1) and PCA-based analysis (Figure 2). Both are used to identify loci under selection, but they differ in their underlying methods. FST is a standardized measure of the variance of allele frequencies among populations [(@wright1949population)](#refs). By directly comparing genetic differentiation between populations/groups to neutral expectations, FST outlier tests can identify loci which may be under selection (Figure 1). In contrast, PCA-based methods use Principal Component Analysis (PCA) to capture population structure and can identify loci with extreme loadings on principal components as potential outliers (Figure 2).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 1: Example of an FST-based analysis using OutFLANK. In this dataset, 1000 individuals were collected from across 39 populations  spanning an environmental gradient. Each dot is a SNP and its location in the genome plotted on the X-axis. The Y-axis is the FST for each SNP. Higher FST values are indicative of larger genetic differentiation and candidate SNPs under selection are highlighted in pink. [Source](https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html)"}
knitr::include_graphics("images/Fig1.png")

@@

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 2: Example of a PCA-based analysis using pcadapt. In this dataset, a total of 150 individuals from three different populations were genotyped at 1,500 markers. Population structure was identified for three discrete groups of individuals, matching population assignment expectations (A). SNPs that are significantly associated with PC1 and PC2 are given higher weightings and are used as input for a Manhattan plot, where the SNP location in the genome is plotted on the X-axis and the weightings (or -log pvalue) are plotted on the Y-axis. SNPs that have higher weightings will have higher -log pvalues and are candidate SNPs under selection among populations (B). [Source](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html)"}
knitr::include_graphics("images/Fig2.png")

@@

Nonselective evolutionary forces can affect the distribution of outlier loci from loci across a genome [(@lotterhos2014evaluation)](#refs), therefore, the first step is to identify the  underlying distribution that is expected under neutral processes by answering the question;  how much variation in among loci would be expected (i.e. the null distribution of FST) in the absence of selection?  One must first create a null distribution using a set of SNPs (single nucleotide polymorphisms)  that accurately model the distribution expected under neutrality. An accurate null distribution typically requires some filtering of a set of SNPs, and in many cases, the best set of SNPs for neutral parameterization is not the entire set of SNPs that you want to evaluate for outliers. Common filtering steps could include a minor allele frequency/count, removing missing data and related individuals, genotype depth and quality, and removing SNPs in linkage disequilibrium (LD). As a general guide, we recommend the following [SNP filtering tutorial](https://ddocent.com//filtering/). Deciding on how much filtering is required can be tricky and can feel subjective. We recommend visualizing the filtering steps using a PCA, which will elucidate many common biases in the dataset (see the [MarineOmics page on PCAs](https://https://marineomics.github.io/POP_08_PCA.html)). We recommend using the same set of SNPs that you use for estimating population structure for neutral parameterization (which is often a set of SNPs not in LD). Once neutral parameterization is performed on one set of SNPs, for most programs that model can be applied to the larger filtered SNP dataset (that across the genome for evaluating their outlier status.

### Step 2 for detecting outlier loci - choose an outlier method based on the biology of your system!

When choosing an outlier method, it is important to understand what biological assumptions are being ignored/violated and whether the model accounts for the demographic history when estimating divergence among populations (see section “Considerations” above). For example, early models such as fdist and fdist2 [(@beaumont1996evaluating)](#refs) assume populations are all equal in size and there is equal probability of migration and gene flow among all populations (even if some populations are located further away than others). This method produced a lot of false positive outliers because more realistic scenarios violated these simple assumptions [(@lotterhos2014evaluation)](#refs). An improvement over this approach was implemented in the program BayeScan where each population has its own neutral model [(@foll2008genome)](#refs), but like fdist/fdist2, BayeScan also  suffers from a large false positive rate due to its violated assumptions of non-independence among populations. Newer outlier methods attempt to account for population demographic history and independence in different ways. For example,  pcadapt uses PC loadings (insead of Fst distributions) which are meant to be less reliant on assumptions about the underlying demography [(@duforet2015haplopop)](#refs). However, PC loadings can be influenced by recombination variation, so it is important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs). These estimates of population structure are included as covariates in the final Outlier Analysis model. See Table 1 below for more information on how each method accounts for population demographic history.

Population structure is also important to consider. For example, if you discover only one genetic cluster (K=1 in STRUCTURE or Admixture plot), perhaps any of the methods in Table 1 would work well for your data. On the other hand, if there is a lot of population structure (i.e., K>1), then a method that accounts for demographic history should be chosen. Additionally, consider what assumptions are made in the tutorials that you use as a resource. Does the tutorial you are using assume that your data is normally distributed? Multiple tutorials have been added below to help with comparing between different assumptions made for different kinds of data.

@@

### TABLE 1. Common methods for identifying outliers and the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method.
| Model & Approach                                                                                                                                                                                                                                                                                | Demographic History                             | Assumptions/ Known issues                                                                                                                                              | False Positive Rate                                                                                                                                     | Tutorial                                                                                                                                                                                                                                                   | Citation                                                            |
| -------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------- |

@@

| PCAdapt (PC loadings) | * Latent factors to estimate values of K <br>* uses PC loadings (less reliant on assumptions about the underlying demography) [(@duforet2015haplopop)](#refs) <br>* Important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs)                           | * individuals not sorted into populations ahead of time <br>* Performs well in populations with high admixture                                                         | Low to medium [(@capblancq2018evaluation)](#refs), [(@luu2017pcadapt)](#refs) Might not be suitable for populations with complex demographic histories. | [Luu, Privé and Blum tutorial](https://https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics course tutorial ](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [(@duforet2015haplopop)](#refs)  [(@silliman2019population)](#refs) |
| Outflank (Weir and Cockerham’s Fst)                                                                                                                                                                                                                                                                | * populations defined ahead of time             | * estimates the number of K genetic clusters to inform a null model of Fst distributions <br>* Performs poorly in populations with admixture and pooledseq samples     |                                                                                                                                                         |                                                                                                                                                                                                                                                            |                                                                     |* finite sample sizes in pop                                                                                                                                                                                                                                                                                                                                        | Low, but suffers from low power                                                                                                                         | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html) (2018) [UC Davis Marine Genomics course tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers)  | [(@lotterhos2014evaluation)](#refs) [(@schlotterer2014sequencing)](#refs) |

@@

| BayPass XTX statistics                       | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs)                                                                                                                                            | * Uses a covariance matrix of allele frequencies among populations  [(@pickrell2012inference)](#refs) * can be used with or without environmental data                                                                                                                                                                           | low                                                                                                                                                     |[Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/)                                                                                                                            | [(@gautier2015genome)](#refs)                                             |

@@

The distribution of genetic diversity within natural populations is significantly influenced by environmental conditions. Along environmental gradients, many marine species demonstrate genetic variation for adaptive traits [(@pinsky2021diversification)](#refs). Genotype environment association (GEA) analysis is a statistical framework that allows us to investigate these genotype-environment relationships and identify specific genetic markers (loci) associated with environmental variation across populations [(@rellstab2015practical)](#refs). In a heterogeneous environment, environmental factors may act as selective pressure for local adaptation. In this case, allele frequencies at loci underlying adaptive phenotypic traits should be associated with environmental factors that act as proxies for the unobserved selective pressures [(@frichot2015lea)](#refs). GEAs are distinct from  outlier detection methods previously described as they do not rely on an allele frequency distribution, but instead use the correlation between genotypes and environment for identifying both loci under selection and key climatic factors driving local adaptation. GEA analyses usually control for isolation by distance in addition to demography (i.e. that associations between genotype and environment are not attributed to environments being far away from one another).

## General Method Overview: GEA Methods 
GEA broadly refers to any methodology that assesses the association between allele frequencies and environmental variables. Therefore, you will need four things before getting into GEAs - a good sampling regime, genetic data, geo-referenced environmental variables (see Section x). GEA methods commonly cited in the literature include: Redundancy Analysis (RDA) Gradient Forest (GF), Bayesian methods (i.e. Bayenv) and Latent Factor Mixed Models (LFMM).

@@

## So you want to do a GEA?
So you want to perform GEA? You will need five things before getting started...

@@

## Experimental Design Conditions for GEA
### 1. A good sampling regime 
Sampling design has implications for minimizing false-discoveries and for detecting true positives (statistical power) in relation to environmental selection. Compared with random or opportunistic sampling approaches, informed sampling designs based on the characteristics of the study area and species life history tend to improve the power of GEAs [(@whitlock2015reliable)](#refs) . The general consensus suggests “stratified sampling” [(@dauphin2023re)](#refs) - a sampling regime that is informed by a priori knowledge of neutral genetic differentiation (or geographic variation) and/or environmental representativeness to capture genetic diversity and population structure. To action a stratified sampling method you would collect an equal number of samples from each group, where each group is identified from  environmental and population structure information  ([(@dauphin2023re)](#refs), Figure 7). If you  are unsure of the genetic structure of your species, it is recommended that you conduct a PCA or build a map of the environmental variables across your potential study area, with the aim to visualize the environmental variation across geographical sites before sampling (see section 4 on Environmental data below). In this case, sampling locations that are environmentally different, but replicated within environments can help distinguish selection from neutral population structure. A key problem in this field is when unobserved variables  correlate with both the primary environmental variables and the genomic data. For example, GEAs tend to have low power and high rates of false positives when environmental gradients are correlated with neutral population structure ([(@whitlock2015reliable)](#refs); [(@caye2019lfmm)](#refs)). Several methods have been specifically developed to account for this issue (see Table 2).

@@

### 4. Environmental data
Lastly, you will need geo-referenced environmental data (see Table 3). There are two key types of environmental data that are used; in situ or spatially interpolated data:

@@

Spatially interpolated data are commonly derived from the assimilation of atmospheric and oceanographic data from models, satellites and in situ observations. These ocean models usually provide accurate estimates of the state of the system, and are commonly used for GEAs because they have broad spatial and temporal resolution, and are usually publicly available and accessible. Some of the useful databases hosting past, present and future climate information  can be found in Table 3.

@@

* https://ropensci.org/blog/2019/11/05/tidync/

* https://ouicodedata.com//posts/worldclim-with-r/

* https://jorgemfa.medium.com/easy-to-use-marine-climate-layers-for-ecological-modelling-83950cd7a75f

* https://github.com/bio-oracle/biooracler

* https://www.bio-oracle.org/

* [(@assis2018bio)](#refs)– paper on bioclimatic modelling

>>>>

---
title: "A MarineOmics overview of methods to detect signatures of selection"
author: "Joanna Griffiths, Danielle Davenport, Madison Armstrong, Camille Rumberger"
date: "2025-02-24"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: true
bibliography: "common-bib_01.bib"
---

*Created using [HackMD](https://hackmd.io/) and [tablesgenerator](https://www.tablesgenerator.com/markdown_tables)*

## Target audience
This page is designed for beginners (with some basic undergraduate-level of understanding of evolutionary biology) and provides a broad introduction to different methods commonly used to study natural selection in wild populations at a microevolutionary scale. On this page we will explore how techniques like outlier analyses, genotype-environment analysis (GEA) and genome-wide association studies (GWAS) can help us study natural selection. To learn more about specific methods, read through the suggested readings in each section and visit the tutorials where links are provided.

@@

<ins>On this page we describe three classes of methods to identify loci under selection at a microevolutionary scale using samples collected from a single time point across space/environments. .</ins>
1. **Outlier Analyses** identify extreme deviations from what would be expected under neutral evolution. 
2. **Genotype-Environment-Analysis (GEA)** links genomic variation to environmental gradients.
3. **Genome-Wide Association Studies (GWAS)** examine genetic variants and identify those statistically associated with a particular trait. When a specific phenotype shows a significant association with a particular loci, gene or region, it suggests that this genetic variation may be under selection.

## Which method should I use to find loci under selection?
The method you choose will partly depend on the data that you have available: Outlier analyses require only genetic data, but GEA requires environmental data across sampled populations, and GWAS requires trait data for each genotyped individual. Addtionally, the biology of your system is always important to consider before conducting specific analyses, and no matter which method(s) you use, it is recommended that comparisons of outlier vs non-outliers (neutral loci) are performed [(@riginos2016navigating)](#refs) as a kind-of “validation” step, where we would expect loci under selection (i.e. outliers) to demonstrate a stronger signal of environmental dependence than non-outliers. However, a common but conservative approach taken by researchers performing outlier detection is to apply multiple test methods and identify outlier SNPs that are overlapping across them. For example, some studies perform a differentiation outlier test first (see [Outlier Analyses Section](#outlier-analyses)) and then use only the resulting outliers as input for the Genome-Environment Associations  [(@rellstab2015practical)](#refs).

Importantly, to conduct outlier analyses you do not necessarily need to have neutral genetic differentiation (genetic variation among populations  primarily driven by random processes like genetic drift and migration) in your data. For example, marine invertebrates with pelagic larval dispersal and high gene flow which show very low to no detectable population structure at neutral alleles across their range (i.e using STRUCTURE, ADMIXTURE, PCA) still show signals of adaptation. Other examples include Pacific purple sea urchins ([(@petak2023whole-cp)](#refs);[(@rumberger2023selection)](#refs)) or scallops [(@van2017identifying)](#refs).

@@

Genetic differences among groups may be driven by a combination of genetic drift, demography and/ or selection. Each of these processes can directly change the allele frequency of a population. However, alleles that are under selection and influence fitness are expected to occur at  frequencies different to those affected only by neutral processes. For example, loci under divergent selection (and those physically linked to them) exhibit stronger differentiation than neutral regions and so we might expect these loci to be more differentiated between populations/groups.  We can identify these differences in allele frequencies among different populations/environments using outlier tests. Outlier tests for selection assume a null model of neutrality, where allele frequencies drift randomly without selection pressures. Loci with significantly higher or lower differentiation (i.e. Fst) between populations/groups than expected under a neutral model are flagged as potential targets of selection. Specifically, while selectively neutral loci are expected to form a normal frequency distribution,  when alleles increase fitness (positive selection) they can be expected to be observed at higher frequencies and where they decrease fitness (negative selection) they can be expected to be observed at at lower frequencies [(@hoban2021genetic)](#refs).  These are  “outliers” or “putatively adaptive loci” which may be candidates of selection.

There are two commonly applied approaches to outlier analysis; Fst-based analysis (Figure 1) and PCA-based analysis (Figure 2). Both are used to identify loci under selection, but they differ in their underlying methods. Fst is a standardized measure of the variance of allele frequencies among populations [(@wright1949population)](#refs). By directly comparing genetic differentiation between populations/groups to neutral expectations, Fst outlier tests can identify loci which may be under selection (Figure 1). In contrast, PCA-based methods use Principal Component Analysis (PCA) to capture population structure and can identify loci with extreme loadings on principal components as potential outliers (Figure 2).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 1: Example of an Fst-based analysis using [OutFLANK](https://rstudio-pubs-static.s3.amazonaws.com/305384_9aee1c1046394fb9bd8e449453d72847.html). In this dataset, 1000 individuals were collected from across 39 populations  spanning an environmental gradient. Each dot is a SNP and its location in the genome plotted on the X-axis. The Y-axis is the Fst for each SNP. Higher Fst values are indicative of larger genetic differentiation and candidate SNPs under selection are highlighted in pink. [Source](https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html)"}
knitr::include_graphics("images/Fig1.png")

@@

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 2: Example of a PCA-based analysis using [pcadapt](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html). In this dataset, a total of 150 individuals from three different populations were genotyped at 1,500 markers. Population structure was identified for three discrete groups of individuals, matching population assignment expectations (A). SNPs that are significantly associated with PC1 and PC2 are given higher weightings and are used as input for a Manhattan plot, where the SNP location in the genome is plotted on the X-axis and the weightings (or -log pvalue) are plotted on the Y-axis. SNPs that have higher weightings will have higher -log pvalues and are candidate SNPs under selection among populations (B). [Source](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html)"}
knitr::include_graphics("images/Fig2.png")

@@

Non-selective evolutionary forces can affect the distribution of outlier loci from loci across a genome [(@lotterhos2014evaluation)](#refs), therefore, the first step is to identify the  underlying distribution that is expected under neutral processes by answering the question;  how much variation in among loci would be expected (i.e. the null distribution of Fst) in the absence of selection?  One must first create a null distribution using a set of SNPs (single nucleotide polymorphisms)  that accurately model the distribution expected under neutrality. An accurate null distribution typically requires some filtering of a set of SNPs, and in many cases, the best set of SNPs for neutral parameterization is not the entire set of SNPs that you want to evaluate for outliers. Common filtering steps could include a minor allele frequency/count, removing missing data and related individuals, genotype depth and quality, and removing SNPs in linkage disequilibrium (LD). As a general guide, we recommend the following [SNP filtering tutorial](https://ddocent.com//filtering/). Deciding on how much filtering is required can be tricky and can feel subjective. We recommend visualizing the filtering steps using a PCA, which will elucidate many common biases in the dataset (see the [MarineOmics page on PCAs](https://marineomics.github.io/POP_08_PCA.html)). We recommend using the same set of SNPs that you use for estimating population structure for neutral parameterization (which is often a set of SNPs not in LD). Once neutral parameterization is performed on one set of SNPs, for most programs that model can be applied to the larger filtered SNP dataset.

### Step 2 - choose an outlier method based on the biology of your system!

When choosing an outlier method, it is important to understand what biological assumptions are being ignored/violated and whether the model accounts for the demographic history when estimating divergence among populations (see [Table 1](#Table1)). For example, early models such as fdist and fdist2 [(@beaumont1996evaluating)](#refs) assume populations are all equal in size and there is equal probability of migration and gene flow among all populations (even if some populations are located further away than others). This method produced a lot of false positive outliers because more realistic scenarios violated these simple assumptions [(@lotterhos2014evaluation)](#refs). An improvement over this approach was implemented in the program BayeScan where each population has its own neutral model [(@foll2008genome)](#refs), but like fdist/fdist2, BayeScan also  suffers from a large false positive rate due to its violated assumptions of non-independence among populations. Newer outlier methods attempt to account for population demographic history and independence in different ways. For example,  pcadapt uses PC loadings (insead of Fst distributions) which are meant to be less reliant on assumptions about the underlying demography [(@duforet2015haplopop)](#refs). However, PC loadings can be influenced by recombination variation, so it is important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs). These estimates of population structure are included as covariates in the final Outlier Analysis model. See [Table 1](#Table1) for more information on how each method accounts for population demographic history.

Population structure is also important to consider. For example, if you discover only one genetic cluster (K=1 in STRUCTURE or Admixture plot), perhaps any of the methods in [Table 1](#Table1) would work well for your data. On the other hand, if there is a lot of population structure (i.e., K>1), then a method that accounts for demographic history should be chosen. Additionally, consider what assumptions are made in the tutorials that you use as a resource. Does the tutorial you are using assume that your data is normally distributed? Multiple tutorials have been added below to help with comparing between different assumptions made for different kinds of data.

@@

### TABLE 1. Common methods for identifying outliers {#Table1}
Common methods for identifying outliers and the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method.

| Model & Approach | Demographic History | Assumptions / Known issues | False Positive Rate | Tutorial | Citation |
|------------------|---------------------|-----------------------------|----------------------|----------|----------|

@@

| PCAdapt (PC loadings) | * Latent factors to estimate values of K <br>* uses PC loadings (less reliant on assumptions about the underlying demography) [(@duforet2015haplopop)](#refs) <br>* Important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs)                           | * individuals not sorted into populations ahead of time <br>* Performs well in populations with high admixture                                                         | Low to medium [(@capblancq2018evaluation)](#refs), [(@luu2017pcadapt)](#refs) Might not be suitable for populations with complex demographic histories. | [Luu, Privé and Blum tutorial](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics course tutorial ](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [(@duforet2015haplopop)](#refs)  [(@silliman2019population)](#refs) |
| Outflank (Weir and Cockerham’s Fst) | * populations defined ahead of time | * Estimates the number of K genetic clusters to inform a null model of Fst distributions <br>* Performs poorly in populations with admixture and PoolSeq samples <br>* Affected by finite sample sizes | Low, but suffers from low power | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html) (2018), [UC Davis Marine Genomics course tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers) | [(@lotterhos2014evaluation)](#refs), [(@schlotterer2014sequencing)](#refs) |

@@

| BayPass XTX statistics | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs) | * Uses a covariance matrix of allele frequencies among populations [(@pickrell2012inference)](#refs) <br> * Can be used with or without environmental data | Low | [Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [(@gautier2015genome)](#refs) |

| Model & Approach | Demographic History | Assumptions / Known issues | False Positive Rate | Tutorial | Citation |
|------------------|---------------------|-----------------------------|----------------------|----------|----------|
| fdist & fdist2 (Fst; using Arlequin software) | * Accounts for loci with different heterozygosities and with finite sample sizes per subpopulation | * Assumes populations are equal in size and migration rates  <br> * Non-independence among populations | High [@lotterhos2014evaluation](#refs); [@narum2011comparison](#refs); [@perez2010comparing](#refs); [@whitlock2015reliable](#refs) | Outdated due to high false positive rates | [@beaumont1996evaluating](#refs) |
| Bayescan (Fst) | * Populations defined ahead of time  <br> * Dirichlet-multinomial distribution: each population has its own parameter for neutral model | * Does not allow for non-random correlations among pairs of populations <br> * Slow run time [@luu2017pcadapt](#refs) | Low to high depending on the demography | [Benestan tutorial (2020)](https://rpubs.com/lbenestan/outlier) | [@foll2008genome](#refs) |
| PCAdapt (PC loadings) | * Latent factors estimate values of K <br> * Uses PC loadings; less reliant on assumptions about the underlying demography [@duforet2015haplopop](#refs) <br> * Important to use a set of SNPs thinned for linkage disequilibrium [@lotterhos2019effect](#refs) | * Individuals not sorted into populations ahead of time <br> * Performs well in populations with high admixture | Low to medium [@capblancq2018evaluation](#refs); [@luu2017pcadapt](#refs) <br> Might not be suitable for populations with complex demographic histories | [Luu, Privé and Blum tutorial](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [@duforet2015haplopop](#refs); [@silliman2019population](#refs) |
| Outflank (Weir & Cockerham’s Fst) | * Populations defined ahead of time | * Estimates the number of K clusters to inform a null Fst model <br> * Performs poorly in populations with admixture and PoolSeq <br> * Affected by finite sample sizes | Low, but suffers from low power | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html), [UC Davis tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers) | [@lotterhos2014evaluation](#refs); [@schlotterer2014sequencing](#refs) |
| FLK / TF-LK | * Sensitive to complex migration patterns and sharp differences in migration rates [@bonhomme2010detecting](#refs) | * Uses a population tree (without migration) to describe the covariance among populations to build a null Fst model | Low [@bonhomme2010detecting](#refs) | [UoF Bioinformatics Hub Tutorial](https://uofabioinformaticshub.github.io/GumCreek2012/scripts/S1_FLK) | See Supplementary File 1 from [@bonhomme2010detecting](#refs) |
| BayPass (XTX statistics) | * Assumes HWE for PoolSeq data [@gautier2015genome](#refs) | * Uses a covariance matrix of allele frequencies among populations [@pickrell2012inference](#refs) <br> * Can be used with or without environmental data | Low | [Barley Agroclimatic Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielsen Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [@gautier2015genome](#refs) |

@@

The distribution of genetic diversity within natural populations is significantly influenced by environmental conditions. Along environmental gradients, many marine species demonstrate genetic variation for adaptive traits [(@pinsky2021diversification)](#refs). Genotype environment association (GEA) analysis is a statistical framework that allows us to investigate these genotype-environment relationships and identify specific genetic markers (loci) associated with environmental variation across populations [(@rellstab2015practical)](#refs). In a heterogeneous environment, environmental factors may act as selective pressure for local adaptation. In this case, allele frequencies at loci underlying adaptive phenotypic traits should be associated with environmental factors that act as proxies for the unobserved selective pressures [(@frichot2015lea)](#refs). GEAs are distinct from  outlier detection methods previously described as they do not rely on an allele frequency distribution, but instead use the correlation between genotypes and environment for identifying both loci under selection and key climatic factors driving local adaptation. GEA analyses usually control for isolation by distance in addition to demography (i.e. ensuring that associations between genotype and environment are not driven by physical distance between environments).

## General Method Overview: GEA Methods 
GEA broadly refers to any methodology that assesses the association between allele frequencies and environmental variables. Therefore, you will need four things before getting into GEAs - a good sampling regime, genetic data, geo-referenced environmental variables (see [Section 4. Environmental data](#env-data)). GEA methods commonly cited in the literature include: Redundancy Analysis (RDA), Gradient Forest (GF), Bayesian methods (i.e. Bayenv), and Latent Factor Mixed Models (LFMM).

@@

## So you want to do a GEA?
So you want to perform GEA? You will need four things before getting started...

@@

## Experimental Design Conditions for GEA
### 1. A good sampling regime 
Sampling design has implications for minimizing false-discoveries and for detecting true positives (statistical power) in relation to environmental selection. Compared with random or opportunistic sampling approaches, informed sampling designs based on the characteristics of the study area and species life history tend to improve the power of GEAs [(@whitlock2015reliable)](#refs) . The general consensus suggests “stratified sampling” [(@dauphin2023re)](#refs) - a sampling regime that is informed by a priori knowledge of neutral genetic differentiation (or geographic variation) and/or environmental representativeness to capture genetic diversity and population structure. To conduct a stratified sampling method you would collect an equal number of samples from each group, where each group is identified from  environmental and population structure information  ([(@dauphin2023re)](#refs), Figure 7). If you  are unsure of the genetic structure of your species, it is recommended that you conduct a PCA or build a map of the environmental variables across your potential study area, with the aim to visualize the environmental variation across geographical sites before sampling (see [section 4 on Environmental data](#env-data)). In this case, sampling locations that are environmentally different, but replicated within environments can help distinguish selection from neutral population structure. A key problem in this field is when unobserved variables  correlate with both the primary environmental variables and the genomic data. For example, GEAs tend to have low power and high rates of false positives when environmental gradients are correlated with neutral population structure ([(@whitlock2015reliable)](#refs); [(@caye2019lfmm)](#refs)). Several methods have been specifically developed to account for this issue (see [Table 2](#Table2)).

@@

### 4. Environmental data {#env-data}
Lastly, you will need geo-referenced environmental data (see [Table 3A](#Table3A) and [Table 3B](#Table3B)). There are two key types of environmental data that are used; in situ or spatially interpolated data:

@@

Spatially interpolated data are commonly derived from the assimilation of atmospheric and oceanographic data from models, satellites and in situ observations. These ocean models usually provide accurate estimates of the state of the system, and are commonly used for GEAs because they have broad spatial and temporal resolution, and are usually publicly available and accessible. Some of the useful databases hosting past, present and future climate information  can be found in [Table 3A](#Table3A), [Table 3B](#Table3B), and [Table 3C](#Table3C).

@@

* [Data extraction from the NetCDF databases](https://ropensci.org/blog/2019/11/05/tidync/)

* [How to extract and plot data from Worldclim](https://ouicodedata.com//posts/worldclim-with-r/)

* [Data extraction from the Bio-ORACLE databse](https://jorgemfa.medium.com/easy-to-use-marine-climate-layers-for-ecological-modelling-83950cd7a75f)

* [Another data extraction method from the Bio-ORACLE database](https://github.com/bio-oracle/biooracler)

<ins>Further recommended reading:</ins>
* [(@rellstab2015practical)](#refs) –method comparison
* [(@lotterhos2015relative)](#refs)–method comparison
* [(@franccois2016controlling)](#refs)–method comparison for false discovery rates
* [(@forester2016detecting)](#refs) (GEA)-- method comparison
* [(@dauphin2023re)](#refs)–discussion of how different types/scales of of environmental data can be used to answer different questions
* [(@capblancq2018evaluation)](#refs) -- method comparison (GEA & pcadapt)
* [(@alvarado2022genotype)](#refs) - example paper that uses gradientForest
* [(@riginos2016navigating)](#refs) – spatial scale in landscape genomics
* [(@capblancq2021redundancy)](#refs) –RDA
* [(@selmoni2020seascape)](#refs)– example landscape genomics study
* [(@assis2018bio)](#refs)– paper on bioclimatic modelling

### TABLE 2: Common methods for identifying candidate SNPs associated with environmental data (GEA) {#Table2}
Common methods for identifying candidate SNPs associated with environmental data (GEA), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method.
